---
title: "2. scrnaseq  GSE164772 data_monocle3"
author: "Lai, Kei Onn (Barron Lab)"
date: "1/5/2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SeuratWrappers)
library(Seurat)
library(monocle3)
library(SingleCellExperiment)
```

Here i read in the seurat object (mgclusters_only_seuratint.rds) from the previous analysis(1. Seurat_scrnaseq_GSE164772.Rmd) as "seurat.rds"
seurat.rds thus contains only microglia population from GSE164772.
Previously, using seurat,i identified 7 microglia populations.
In this pipeline, we will use monocle3 for a trajectory anlaysis for these 7 microglia
population in zebrafish.


creating cell dataset object from seurat object using SeuratWrapper
```{r}
cds<-as.cell_data_set(seurat)
#meta data
colData(cds)
```
getting feature data aka the data of genes
```{r}
fData(cds)
fData(cds)$gene_short_name<-rownames(fData(cds))
fData(cds)
# we added extra col of gene_short_name

#getting counts
counts(cds)[1:6,1:6]
```


For trajectory analysis, monocle3 needs cell embeddings of UMAP, partitions and the clusters. so, i used cluster_cells() from monocle3 to these information rather than using just seurat object.

```{r}
cds <- cluster_cells(cds, resolution=1e-3)

p1 <- plot_cells(cds, color_cells_by = "cluster", show_trajectory_graph = FALSE)
p2 <- plot_cells(cds, color_cells_by = "partition", show_trajectory_graph = FALSE)
p1| p2

integrated.sub <- subset(as.Seurat(cds, assay = NULL), monocle3_partitions == 1)
cds <- as.cell_data_set(integrated.sub)
cds <- learn_graph(cds, use_partition = TRUE, verbose = FALSE)

trajplot<-plot_cells(cds,
           color_cells_by = "cluster",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE)+theme(legend.position = "right")
```


```{r trajplot}


trajplot
```


we know that microglia population 6 is proliferative microglia and likely to be the
root of the pseudotime plot 
```{r}
cds <- order_cells(cds, root_cells = colnames(cds[,clusters(cds) == 6]))
pseudot<-plot_cells(cds,
           color_cells_by = "pseudotime",
           group_cells_by = "cluster",
           label_cell_groups = FALSE,
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           label_roots = FALSE,
           trajectory_graph_color = "grey60")
pseudot
```
Since cluster 3 is furthest from cluster 6,
we do pairwise comparison

```{r}
cluster3vs6<-FindMarkers(seurat, ident.1 = "Microglia3", ident.2 = "Microglia6")
head(cluster3vs6)

```

extracting genes that vary across the pseudotime
```{r ,eval=F}
cds_graph_test_results <- graph_test(cds,
                                     neighbor_graph = "principal_graph",
                                     cores = 8)
#write.csv(cds_graph_test_results,"genes_vary_pseudotime.csv")

rowData(cds)$gene_short_name <- row.names(rowData(cds))
head(cds_graph_test_results, error=FALSE, message=FALSE, warning=FALSE)

deg_ids <- rownames(subset(cds_graph_test_results[order(cds_graph_test_results$morans_I, decreasing = TRUE),], q_value < 0.05))

topgenes_overlay_pseudo<-plot_cells(cds,
           genes=head(deg_ids)),
           show_trajectory_graph = F,
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           scale_to_range=T,
           norm_method = "log")
```

plotting as a function of time,
in this case, we are interested in the top few genes of deg_ids


```{r}

top_genes <- head(deg_ids)

plot_pseudot<-plot_genes_in_pseudotime(top_genes,color_cells_by = "monocle3_clusters"
                                   ,cell_size = 5)
plot_pseudot
```


